---
title: Identification of robust and abundant reference transcripts for EV mRNA cargo normalization
author: Radosław Grochowski
affiliation: Karolinska Institutet, Stockholm
date: 2024
---

# Identification of robust and abundant reference transcripts for EV mRNA cargo normalization

## Description

Following notebook contains code used for RNA-seq data analysis and visualization
for Zickler et al. (2025). RPKM values are calculated and ranked for EV samples 
of various origin.

## Code
### Setup
```{r setup}
library(edgeR) # for RPKM calculation
library(GenomicFeatures) # used for gene length calculation from GTF files
library(annotables) # Ensembl id to symbol conversion
library(dplyr) # data handling
library(data.table) # data handling
library(Matrix) # data handling
library(ggplot2) # visualization
library(ggpubr) # visualization
library(ggnewscale) # visualization

# set the theme for plotting
theme_set(new = theme_minimal())
theme_update(
    strip.background = element_rect(fill = "#FFFFFF"),
    plot.title = element_text(hjust = 0.5, size = 19),
    plot.subtitle = element_text(size = 13, hjust = .5),
    axis.title = element_text(size = 15),
    axis.text.y = element_text(size = 13, face = "bold"),
    axis.text.x = element_text(size = 12, vjust = 0.5),
    legend.text = element_text(size = 13),
    legend.key.size = unit(2, "line"),
    legend.title = element_text(size = 15, hjust = .5, vjust = .5)
)

# That's not necessary (rmarkdown sets its 
# working directory to where the .Rmd file is.)
wd <- "working directory"
knitr::opts_knit$set(root.dir = wd)

# Lapply functions

# calculate min RPKM setting the value
# to 0 if any of the replicates is 0
calc_mean_rpkm_w0 <- function(group, dt_rpkm) {
    dt_rpkm <- as.data.table(dt_rpkm)
    rpkm <- dt_rpkm[, colnames(dt_rpkm) %like% group, with = FALSE]
    # replace rows with at least one 0 with 0s
    rpkm <- apply(rpkm, 1, function(x) {
        x[any(x == 0)] <- 0
        return(x)
    })
    # transpose after apply
    rpkm <- t(rpkm)
    # calculate means
    matrix <- matrix(rowSums(rpkm) / dim(rpkm)[2])
    colnames(matrix) <- group
    return(matrix)
}
# without zero removal for calculations for plasma samples
calc_mean_rpkm <- function(group, dt_rpkm) {
    dt_rpkm <- as.data.table(dt_rpkm)
    rpkm <- dt_rpkm[, colnames(dt_rpkm) %like% group, with = F]
    # calculate means
    matrix <- matrix(rowSums(rpkm)/dim(rpkm)[2])
    colnames(matrix) <- group
    return(matrix)
}

# a function sorting genes in a column and returning
# a data.table with gene and sample names
rank_genes <- function(sample, dt_rpkm, gene) {
    # set a dt
    dt <- data.table("rpkm" = dt_rpkm[, sample])
    # add gene ids
    dt$gene_id <- gene[, gene_id]
    # add gene names (symbols)
    dt$symbol <- gene[, symbol]
    # add sample
    dt[, sample := sample]
    # sum the expression per symbol
    dt[, rpkm := sum(rpkm), by = .(symbol)]
    # remove duplicated symbols
    dt <- unique(dt, by = "symbol")
    # sort the genes
    dt <- dt[order(rpkm, decreasing = T)]
    # rank the gens
    dt$rank <- 1:nrow(dt)
    # set the min rank to the min of the first 0 gene
    dt[rpkm == 0, rank := min(rank)]

    return(dt)
}


# Gene names for conversion of Ensembl ids
grch37_gene <- as.data.table(grch37[, c("ensgene", "symbol")])
grch38_gene <- as.data.table(grch38[, c("ensgene", "symbol")])
colnames(grch37_gene)[1] <- "gene_id"
colnames(grch38_gene)[1] <- "gene_id"
```


```{r}
# Gene lengths
# GRCh38
gtfFile38 <- "Homo_sapiens.GRCh38.111.gtf"
txdb38 <- makeTxDbFromGFF(gtfFile38, format = "gtf")
exonic38 <- exonsBy(txdb38, by = "gene")
red_exonic38 <- reduce(exonic38)
exon_lengths38 <- data.frame("exon_lengths" = vapply(
    width(red_exonic38),
    sum, numeric(1)
))
exon_lengths38$gene_id <- rownames(exon_lengths38)
# GRCh37
gtfFile37 <- "Homo_sapiens.GRCh37.87.gtf"
txdb37 <- makeTxDbFromGFF(gtfFile37, format = "gtf")
exonic37 <- exonsBy(txdb37, by = "gene")
red_exonic37 <- reduce(exonic37)
exon_lengths37 <- data.frame("exon_lengths" = vapply(
    width(red_exonic37),
    sum, numeric(1)
))
exon_lengths37$gene_id <- rownames(exon_lengths37)
```


### Cultured media EVs

#### Gene ranking
```{r}
# load the data
cell_rpkm <- fread("cell_culture_ev_rpkm.csv",
    drop = 1
)
cell_gene <- cell_rpkm[, 1]
cell_rpkm <- cell_rpkm[, -1]
# make a dt with gene names
gt <- data.table(cell_gene)
gt_join <- grch37_gene[gt, on = .(gene_id), mult = "first"][
    is.na(symbol) | symbol == "",
    symbol := gene_id
] # retain unique ENSEMBL ids
# get the cell types
cell_types <- unique(gsub(
    "_[1-3]", "",
    colnames(cell_rpkm)
))
# calculate mean rpkm values per cell type
cell_mean_rpkm <- sapply(cell_types,
    calc_mean_rpkm_w0,
    dt_rpkm = cell_rpkm
)
# rank genes
cell_mean_rank <- lapply(cell_types,
    rank_genes,
    dt_rpkm = cell_mean_rpkm,
    gene = gt_join
)
# turn the list into a data.table
cell_mean_rank <- bind_rows(cell_mean_rank)
setnames(cell_mean_rank, "rpkm", "mean_rpkm")
# set cell type order
cell_type_order <- rev(c(
    "BM-MSC", "CB-MSC", "BJ-5ta", "WJ-MSC",
    "HEK293-T", "HEK293-FS", "CAP", "HAEC",
    "HUVEC", "PANC-1", "THP-1", "JURKAT"
))
cell_mean_rank[, cell_type := factor(sample, levels = cell_type_order)]
setkey(cell_mean_rank, sample)
# calculate rank sums
cell_rank_sum <- cell_mean_rank[, sum(rank), by = .(gene_id, symbol)]
# order the genes
cell_order_symbol <- cell_rank_sum[order(V1), symbol]
# remove those that repeat
cell_order_symbol <- unique(cell_order_symbol)
```
#### Plotting

##### Fig.1C

```{r fig.width=14.4, fig.height=7.5}
# color-code cell types
y_labels_color <- rev(c(
    "green3", "green3", "green3", "green3",
    "blue", "blue",
    "gray35", "gray35",
    "red", "red",
    "orange", "orange"
))
# color chosen gene names
x_labels_color <- ifelse(cell_order_symbol[1:50] %in% c(
    "MT-RNR2", "RP11-217O12.1", "ACTB",
    "MALAT1", "FTL", "GAPDH",
    "BCYRN1", "TMSB4X", "FN1",
    "VIM", "BEST1"
),
"red", "grey30" # default color is grey30
) 
# change font face of chosen gene names
x_labels_face <- ifelse(cell_order_symbol[1:50] %in% c(
    "MT-RNR2", "RP11-217O12.1", "ACTB",
    "MALAT1", "FTL", "GAPDH",
    "BCYRN1", "TMSB4X", "FN1",
    "VIM", "BEST1"
),
"bold", "plain"
)

ggplot(cell_mean_rank[symbol %in% cell_order_symbol[1:50] & mean_rpkm > 1000]) +
    geom_tile(
        data = cell_mean_rank[symbol %in% cell_order_symbol[1:50]],
        aes(
            x = factor(symbol, levels = cell_order_symbol),
            y = factor(cell_type, levels = cell_type_order),
            fill = "RPKM < 1000"
        ), linetype = 0
    ) +
    scale_fill_manual(values = "gray55") +
    labs(fill = "") +
    new_scale_fill() +
    geom_tile(aes(
        x = factor(symbol, levels = cell_order_symbol),
        y = factor(cell_type, levels = cell_type_order),
        fill = mean_rpkm
    )) +
    labs(fill = "Mean RPKM") +
    scale_fill_gradientn(trans = "log10", colours = c("blue", "white", "red")) +
    ggtitle("Top 50 genes above the 1000 RPKM threshold across cell line EV samples") +
    scale_x_discrete(guide = guide_axis(angle = 45)) +
    theme(
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.title = element_text(hjust = .5),
        axis.text.y = ggtext::element_markdown(
            hjust = 1,
            colour = y_labels_color
        ),
        axis.text.x = ggtext::element_markdown(
            hjust = 1,
            colour = x_labels_color,
            face = x_labels_face
        )
    )
```

##### Supplementary Fig.S2A

```{r, fig.width=14, fig.height=7.5}
ggplot(cell_mean_rank[symbol %in% cell_order_symbol[1:50]]) +
    geom_tile(aes(
        x = factor(symbol, levels = cell_order_symbol[1:50]),
        y = factor(cell_type, levels = cell_type_order),
        fill = mean_rpkm
    )) +
    scale_fill_gradientn(trans = "log10", colours = c("blue", "white", "red")) +
    ggtitle("Top 50 genes across cell line EV samples") +
    labs(x = "gene id", fill = "Mean RPKM") +
    scale_x_discrete(guide = guide_axis(angle = 45)) +
    theme(
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = ggtext::element_markdown(
            hjust = 1,
            colour = y_labels_color
        )
    )
```

##### Supplementary Fig.S5A

```{r fig.width=7, fig.height=5}
order_chosen_genes <- which(cell_order_symbol %in% c(
    "ACTB", "FTL",
    "GAPDH", "TMSB4X",
    "FN1", "VIM",
    "BEST1", "RAB13"
))

ggplot(cell_mean_rank[symbol %in% cell_order_symbol[order_chosen_genes]]) +
    geom_boxplot(aes(
        x = factor(symbol, levels = cell_order_symbol),
        y = rank
    )) +
    geom_point(aes(
        x = factor(symbol, levels = cell_order_symbol),
        y = rank
    )) +
    scale_y_log10() +
    ggtitle("Rank spread of chosen genes\nacross cell line EV samples") +
    theme(
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.title = element_text(hjust = .5),
        plot.margin = unit(c(0, 0, 0, .5), "cm")
    )
```

## Biological fluids EVs (without plasma)

```{r}
flu <- fread("biofluids_ev_counts.csv", drop = 1)
flu_gene <- flu[, 1]
flu <- flu[, -1]
# make a dt with gene names
gt <- data.table(flu_gene)
gt_join <- grch38_gene[gt, on = .(gene_id), mult = "first"
                       ][is.na(symbol) | symbol == "", symbol := gene_id]
# match the gene length with the counts
flu_exon_lengths <- exon_lengths38[match(
    flu_gene$gene_id,
    exon_lengths38$gene_id
), ]
# calculate RPKM values
flu_rpkm <- rpkm(flu, gene.length = flu_exon_lengths$exon_lengths)
# rank genes
flu_rank <- lapply(colnames(flu_rpkm),
    rank_genes,
    dt_rpkm = flu_rpkm,
    gene = gt_join
)
# turn the list into a data.table
flu_rank <- bind_rows(flu_rank)
# add a column with sample type
flu_rank[, sample_type := gsub(".*_", "", sample)]
# add a column with donor and format it
flu_rank[, donor := regmatches(sample, regexpr("donor_[1-9]", sample))
         ][, donor := gsub("_", " ", donor)
           ][, donor := gsub("d", "D", donor)]
# calculate rank sums per sample type
flu_rank_sum <- flu_rank[, sum(rank), by = .(gene_id, symbol, sample_type)]
# order_id_flu <- flu_rank_sum[order(V1), gene_id, by = sample_type]
flu_order_symbol <- flu_rank_sum[order(V1), symbol, by = sample_type]
# remove those that repeat
flu_order_symbol <- flu_order_symbol[, symbol := .SD[, unique(symbol)],
    by = sample_type
]
```

#### Fig.5E (urine EVs) and Supplementary Fig.S7B-D

```{r, fig.width=14, fig.height=5}
# set the donor order
flu_donor <- rev(c(
    "Donor 1", "Donor 2", "Donor 3", "Donor 4",
    "Donor 5", "Donor 6", "Donor 7", "Donor 8"
))

plots <- lapply(unique(flu_rank$sample_type), function(x) {
    # color-code chosen genes
    x_labels_color <- ifelse(
        flu_order_symbol[sample_type == x
                       ][1:50, symbol] %in% c("GAPDH", "ACTB", 
                                              "TMSB4X", "VIM", "FTL"), 
        "red", "black")

    ggplot(flu_rank[sample_type == x & 
                      symbol %in% flu_order_symbol[sample_type == x
                                                   ][1:50, symbol]]) +
        geom_tile(aes(
            x = factor(symbol, 
                       levels = flu_order_symbol[sample_type == x
                                                 ][1:50, symbol]),
            y = factor(donor, levels = flu_donor),
            fill = rpkm
        )) +
        scale_fill_gradientn(trans = "log10", 
                             colours = c("blue", "white", "red")) +
        ggtitle(paste0("Top 50 genes across ", x, " EV samples")) +
        labs(x = "gene id", fill = "Mean RPKM") +
        scale_x_discrete(guide = guide_axis(angle = 45)) +
        theme(
            panel.grid.major = element_blank(),
            panel.grid.minor = element_blank(),
            axis.line = element_blank(),
            axis.title.x = element_blank(),
            axis.title.y = element_blank(),
            axis.text.x = ggtext::element_markdown(
                hjust = 1,
                colour = x_labels_color
            ),
            plot.margin = margin(.5, .5, .5, .5, "cm")
        )
})
plots
```

## Plasma EVs

```{r}
plas <- fread("plasma_ev_counts.csv",
    drop = 1
)
plas_genes <- plas[, 1]
plas <- plas[, -1]
# make a dt with gene names
gt <- data.table(plas_genes)
gt_join <- grch38_gene[gt, on = .(gene_id), mult = "first"
                       ][is.na(symbol) | symbol == "", symbol := gene_id]
# match the gene length with the counts
plas_exon_lengths <- exon_lengths38[match(
    plas_genes$gene_id,
    exon_lengths38$gene_id
), ]
# calculate RPKM values
plas_rpkm <- rpkm(plas, gene.length = plas_exon_lengths$exon_lengths)
# match gene lengths with the counts
exon_lengths <- exon_lengths38[match(
    unlist(plas_genes, use.names = F),
    exon_lengths38$gene_id
), ]
plas_rpkm <- rpkm(plas, gene.length = exon_lengths$exon_lengths)
# get the cell types
plas_donor <- unique(gsub(
    "_[1-3]$", "",
    colnames(plas_rpkm)
))
# calculate mean rpkm values per donor
plas_mean_rpkm <- sapply(plas_donor,
    calc_mean_rpkm,
    dt_rpkm = plas_rpkm
)
# rank genes
plas_mean_rank <- lapply(colnames(plas_mean_rpkm),
    rank_genes,
    dt_rpkm = plas_mean_rpkm,
    gene = gt_join
)
# turn the list into a data.table
plas_mean_rank <- bind_rows(plas_mean_rank)
# add a column with donor and format it
plas_mean_rank[, donor := regmatches(sample, regexpr("donor_[1-9]", sample))
               ][, donor := gsub("_", " ", donor)
                 ][, donor := gsub("d", "D", donor)]
# add 1 to rpkm values to allow log10 transformation
# calculate rank sums
plas_rank_sum <- plas_mean_rank[, sum(rank), by = .(gene_id, symbol)]
plas_order_symbol <- plas_rank_sum[order(V1), symbol]
# remove those that repeat
plas_order_symbol <- unique(plas_order_symbol)
```

#### Fig.5D (plasma EVs)

```{r, fig.width=14, fig.height=5}
x_labels_color <- ifelse(plas_order_symbol[1:50] %in% c("GAPDH", "ACTB", 
                                                        "TMSB4X", "VIM", 
                                                        "FTL"), "red", "black")
# set donor order
plas_donor <- rev(c(
    "Donor 1", "Donor 2", "Donor 3",
    "Donor 4", "Donor 5"
))

ggplot(plas_mean_rank[symbol %in% plas_order_symbol[1:50]]) +
    geom_tile(aes(
        x = factor(symbol, levels = plas_order_symbol[1:50]),
        y = factor(donor, levels = plas_donor), fill = rpkm
    )) +
    scale_fill_gradientn(trans = "log10", colours = c("blue", "white", "red")) +
    ggtitle("Top 50 genes across plasma EV samples") +
    labs(x = "gene id", fill = "Mean RPKM") +
    scale_x_discrete(guide = guide_axis(angle = 45)) +
    theme(
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = ggtext::element_markdown(
            hjust = 1,
            colour = x_labels_color
        )
    )
```
